---
title: "WTI Options Portfolio: Full Greek P&L Attribution"
subtitle: "Individual Project"
date: "`r Sys.Date()`"
author: "Moe Moughnieh"
format:
  html:
    theme: cosmo
    embed-resources: true
    page-layout: full
    toc: true
    toc-location: left
    grid:
      margin-width: 200px
      body-width: 1000px
    resource:
      - "*.js"
      - "*.mp3"
editor: source
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE)
options(scipen = 999)
library(tidyverse)
library(gt)
library(RTL)
library(gridExtra)
library(patchwork)
library(scales)

F_0  <- 73.00
r    <- 0.04
mult <- 1000

port <- dplyr::tribble(
  ~pos, ~type,   ~K,     ~T2M,   ~sigma,  ~n,
  "P1", "put",   70.00,  3/12,   0.3500,  +2000L,
  "P2", "put",   55.00,  3/12,   0.4200,  -2000L,
  "P3", "call",  80.00,  3/12,   0.3000,  -2000L
)
```

# Description

You are a risk analyst at an energy trading desk. The desk holds a portfolio of WTI crude oil options — specifically, options on NYMEX CL futures contracts.

Your role is to:

-   Build a full-Greek P&L attribution framework: the standard tool used on trading desks to explain daily P&L movements across **Delta** (directional exposure), **Gamma** (curvature), **Vega** (volatility sensitivity), **Vanna** (sensitivity of delta to volatility), and **Theta** (time decay).
-   You are expected to understand contracts and use the right option model and specifications in the `RTL` package.
-   Tell me in your own words the portfolio's risk profile, how the Greeks interact, and what this means for a desk managing this book in practice.
-   No AI use in any shape or form and assignment submitted should be a github repo that I can clone and assess.

# Objective

1.  Compute numerical Greeks for each position and for the overall portfolio across the full WTI price range (\$50–\$100/bbl).
2.  Interpret the portfolio's mark-to-market profile and explain how changing Gamma across strike levels affects delta-hedging in practice.
3.  Build a complete P&L attribution framework — Delta, Gamma, Vega, Vanna, and Theta — and verify it accurately explains overnight P&L under a realistic market move.
4.  Critically evaluate where and why the linear/quadratic approximation breaks down as spot moves away from current market.

# Exposure

```{r exposure_table, echo=FALSE}
dplyr::tribble(
  ~Position, ~Type,  ~`Strike K ($/bbl)`, ~Maturity, ~`Impl. Vol`, ~Contracts,
  "P1", "Put",  70.00, "3M", "35.00%", "+2,000 long",
  "P2", "Put",  55.00, "3M", "42.00%", "-2,000 short",
  "P3", "Call", 80.00, "3M", "30.00%", "-2,000 short"
) |>
  gt::gt() |>
  gt::tab_header(
    title    = "WTI Options Portfolio",
    subtitle = "Three-way collar: long put spread + short OTM call  |  F\u2080 = $73/bbl"
  ) |>
  gt::fmt_number(columns = `Strike K ($/bbl)`, decimals = 2)
```

**Market parameters**: current WTI futures price F₀ = **\$73.00/bbl**, risk-free rate r = **4.00%** (annualised, continuously compounded).

Select the appropriate pricing model from the `RTL` package for options on WTI crude oil futures. Consider the nature of the underlying contract — a NYMEX CL futures — when choosing the model and setting its cost-of-carry parameter.

# Question 1 (9 points)

For **each position**, compute the following Greeks **numerically** using central differences evaluated at `F_0`:

-   **Delta** (∂V/∂F)
-   **Gamma** (∂²V/∂F²)
-   **Vega** (∂V/∂σ)
-   **Theta** (∂V/∂T)

**Step size selection**: For each Greek, choose an appropriate finite-difference step size and explain your reasoning. A good step is small enough that the finite difference accurately approximates the true derivative, but not so small that floating-point subtraction errors dominate. Consider the natural scale of each risk factor (price in \$/bbl, vol as a decimal, time in years) and any special considerations for the short-dated options in this portfolio.

**Dollar scaling**: Each raw numerical derivative gives the option price change **per barrel** for a unit move in the risk factor. To express the Greek as a total dollar P&L impact for a position, multiply by `n × mult`, where `n` is the signed contract count (positive = long, negative = short) and `mult = 1,000 bbl/contract`. For example, a delta of 0.50/bbl for a 3,000-contract long position means the desk gains or loses 0.50 × 3,000 × 1,000 = \$1,500,000 per \$1/bbl move in WTI.

Create plots showing each Greek vs. futures price over F ∈ \[50, 100\], with one line per position.

**d.** Compute the **portfolio mark-to-market (MTM)** value and **portfolio Gamma** across F ∈ \[50, 100\]. Plot them on two separate panels. In 3–5 sentences, explain:

-   What the MTM profile shape reveals about the portfolio's net directionality and curvature.
-   Where portfolio Gamma is highest and lowest across the price range, and why those regions correspond to specific positions.
-   What the changing Gamma means for a delta-hedger rebalancing daily as WTI moves through the \$50–\$100 range.

```{r q1}

stepSizeS <- 0.01
stepSizeVol <- 0.01
stepSizeTheta <- 1 / 365

# Customize RTL Option Pricing to given parameters

priceOption <- function(S, X, T2M, r = 0.04, b = 0, sigma, type) {
  RTL::GBSOption(
    S = S,
    X = X,
    T2M = T2M,
    r = r, 
    b = 0,
    sigma = sigma,
    type = type
  )$price
}

port <- port %>%
  dplyr::mutate("Position" = c("LongP @ 70", "ShortP @ 55", "ShortC @ 80")) %>%
  dplyr::select(-pos) %>%
  dplyr::select(Position, everything())

## Construct comprehensive tibble

portTableFull <- tibble(
  underlyingFut = rep(seq(50, 100, by = 1), nrow(port))
) %>%
  dplyr::group_by(underlyingFut) %>%
  dplyr::mutate(Position = port$Position,
                type = port$type,
                K = port$K,
                T2M = port$T2M,
                sigma = port$sigma,
                n = port$n) %>%
  dplyr::arrange(underlyingFut) %>%
  dplyr::ungroup() %>% 
  dplyr::mutate(price = mapply(priceOption, S = underlyingFut, X = K, T2M = T2M, sigma = sigma, type = type),
                priceUpF_0 = mapply(priceOption, S = (underlyingFut + stepSizeS), X = K, T2M = T2M, sigma = sigma, type = type),
                priceDownF_0 = mapply(priceOption, S = (underlyingFut - stepSizeS), X = K, T2M = T2M, sigma = sigma, type = type)) %>%
  dplyr::mutate(delta = n * mult * ((priceUpF_0 - priceDownF_0) / (2 * stepSizeS)),
                gamma = n * mult * ((priceUpF_0 - (2*price) + priceDownF_0) / (stepSizeS^2))) %>%
  dplyr::select(-c(priceUpF_0, priceDownF_0)) %>%
  dplyr::mutate(priceUpVol = mapply(priceOption, S = underlyingFut, X = K, T2M = T2M, sigma = (sigma + stepSizeVol), type = type),
                priceDownVol = mapply(priceOption, S = underlyingFut, X = K, T2M = T2M, sigma = (sigma - stepSizeVol), type = type)) %>%
  dplyr::mutate(vega = n * mult * ((priceUpVol - priceDownVol) / (2*stepSizeVol))) %>% dplyr::select(-c(priceUpVol, priceDownVol)) %>%
  dplyr::mutate(priceTimeDown = mapply(priceOption, S = underlyingFut, X = K, T2M = (T2M - stepSizeTheta), sigma = sigma, type = type)) %>%
  dplyr::mutate(theta = n * mult * ((priceTimeDown - price) / (stepSizeTheta))) %>% dplyr::select(-c(priceTimeDown))
  
##

# Plots of each greek across interval F [50, 100]

# Ideas: Scale y axis, alter x & y names, add title to the joined plots (can I do that?)

greekTable <- portTableFull %>%
  dplyr::select(c(underlyingFut, Position, delta, gamma, vega, theta))

deltaPlot <- greekTable %>%
  ggplot(aes(x = underlyingFut, y = delta, col = Position)) + geom_line(linewidth = 1.5) + labs(x = "Value of F", y = "Delta") + scale_y_continuous(labels = scales::label_number(scale = 0.001, suffix = "K"))

gammaPlot <- greekTable %>%
  ggplot(aes(x = underlyingFut, y = gamma, col = Position)) + geom_line(linewidth = 1.5) + labs(x = "Value of F", y = "Gamma") + scale_y_continuous(labels = scales::label_number(scale = 0.001, suffix = "K"))

vegaPlot <- greekTable %>%
  ggplot(aes(x = underlyingFut, y = vega, col = Position)) + geom_line(linewidth = 1.5) + labs(x = "Value of F", y = "Vega") + scale_y_continuous(labels = scales::label_number(scale = 0.001, suffix = "K"))

thetaPlot <- greekTable %>%
  ggplot(aes(x = underlyingFut, y = theta, col = Position)) + geom_line(linewidth = 1.5) + labs(x = "Value of F", y = "Theta") + scale_y_continuous(labels = scales::label_number(scale = 0.001, suffix = "K"))

##

# d) Portfolio MTM + Portfolio Gamma

# Value of portfolio at F_0 = 73
portValueToday <- portTableFull %>%
  dplyr::filter(underlyingFut == F_0) %>%
  dplyr::mutate(positionVal = n * mult * price) %>%
  dplyr::pull(positionVal) %>%
  sum(.)

# MTM Calculations

portMTM <- portTableFull %>%
  dplyr::select(underlyingFut, Position, n, price, gamma) %>% 
  dplyr::mutate(positionVal = n * mult * price) %>%
  dplyr::group_by(underlyingFut) %>%
  dplyr::summarise(
    portfolioVal = sum(positionVal),
    portfolioGamma = sum(gamma)
  ) %>%
  dplyr::ungroup()

# Plot MTM and Gamma across F [50,100]

portMTMPlot <- portMTM %>%
  ggplot(aes(x = underlyingFut, y = portfolioVal)) + geom_line(linewidth = 1.5, color = "darkblue") + labs(x = "Value of F", y = "Portfolio MTM") + scale_y_continuous(label = scales::label_number(scale = 0.000001, suffix = "M"))

portGammaPlot <- portMTM %>%
  ggplot(aes(x = underlyingFut, y = portfolioGamma)) + geom_line(linewidth = 1.5, color = "darkgreen") + labs(x = "Value of F", y = "Portfolio Gamma")

```

### Step-Size Selection

*Underlying step-size:* \$0.01

*Vega step-size:* 0.01 (1 vol point)

*Theta step-size*: (1 / 365) or 1 calendar day

I decided that these step-sizes were appropriate as they provide ample accuracy while maintaining large enough differences in the numerator so that floating point errors don't dominate. After re-running my function on lower step-sizes, I found that the cost of floating point errors outweighed the benefit of estimation accuracy. I also verified that the greeks are stable across all values of F (by observing the table) for the chosen step-sizes since numerator differences will vary depending on moneyness for each position.

### Greek vs. Futures Price over F ∈ \[50, 100\]

```{r, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}

(deltaPlot | gammaPlot) / (vegaPlot | thetaPlot) + plot_annotation(title = "Greeks vs. WTI Futures over Interval [50,100]", subtitle = "Vertical axis is PnL per 1 unit change in respective risk factor (thousands)")

```

### Portfolio MTM & Portfolio Gamma

```{r, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}

(portMTMPlot | portGammaPlot) + plot_annotation(title = "Portfolio MTM & Gamma over WTI Futures Interval [50,100]")
```

The shape of the portfolio's MTM curve across the interval reveals that it's downward sloping and slightly concave. Portfolio gamma is highest at around \$50, lowest at around \$65, and has a local maximum at around \~80 (in magnitude). In the first interval (50-65) the portfolio gamma decreases in magnitude as we approach the strike price of the long put, which is positive gamma. After its peak, it begins increasing again as futures price approaches the strike of the short call (negative gamma). The gamma of the portfolio then decreases after \~ \$80 as the underlying moves farther away from the strike.

Since gamma is the rate of change of delta, a portfolio manager must rebalance more frequently as the value of WTI futures approaches the strikes of the short positions. The portfolio manager will balance less frequently at values closer to the strike of the long put or beyond the strike of the short call (\$80).

# Question 2 (3 points)

Sum Greeks across positions to show **portfolio-level Greeks** at the current market. Present in a `gt::gt()` table. At approximately what futures price does the portfolio delta change sign?

```{r q2}

# Portfolio greeks @ F_0 = 73

portGreeksToday <- portTableFull %>%
  dplyr::filter(underlyingFut == F_0) %>%
  dplyr::group_by(underlyingFut) %>%
  dplyr::summarise(
    "Delta" = sum(delta),
    "Gamma" = sum(gamma),
    "Vega" = sum(vega),
    "Theta" = sum(theta)
  ) %>%
  dplyr::select(-underlyingFut)


```

```{r, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}

portGreeksToday %>% gt() %>%
  tab_header(title = "Portfolio Greeks Today @ $73.00/bbl WTI Future") %>%
  tab_style(style = cell_borders(sides = "all", color = "black"), locations = cells_body())
```

### Portfolio Delta Sign Change

The sign of the portfolio delta never changes given the positions. This is because at no point across the interval of WTI prices are we ever in a net long delta position. We can prove this by looking at price intervals between and beyond strikes:

**When I say increasing/decreasing, I am referring to magnitude only. Signs are implied by the contract positions.**

-   F \< 55: Delta of short put approaches +1 while delta of the long put approaches -1. Delta of the short call approaches 0. \~ 0 delta for very small values, but expect delta to be slightly negative as it approaches 55 and the puts become deeper in the money.

-   55 \< F \< 70: Delta of the short put begins decreasing as it moves away from strike, delta of the long put increases as it approaches its strike. Net negative delta.

-   70 \< F \< 80: The delta of the long put begins decreasing (becomes less negative), but the delta of the short call becomes more negative as the price approaches 80. Net negative delta.

-   F \> 80: As the price approaches infinity in this interval, the delta for both the long and put approach 0, while the delta for the short call approaches -1. Net negative delta.

# Question 3 (8 points) 

Overnight, US-Iran tensions escalate sharply. The US strikes Iranian nuclear facilities; Iran retaliates by mining the Strait of Hormuz. WTI gaps up at the open:

-   WTI futures: F moves from **73.00** to **90.00** (+\$17.00/bbl)
-   Implied volatility: each position's sigma spikes by **+0.20** (20 vol points)
-   Time: **1 calendar day** passes

For each position and for the portfolio total, compute:

1.  **Actual P&L**: full revaluation at new F, new sigma, new T. P&L = (new price − old price) × n × mult.
2.  **Attributed P&L**: ΔV ≈ (Delta × ΔF) + (½ × Gamma × ΔF²) + (Vega × Δσ) + (Vanna × ΔF × Δσ) + (Theta × Δt), scaled to dollars.
3.  **Unexplained residual**: Actual − Attributed.
4.  **Attribution breakdown**: what share (%) of Attributed P&L comes from each Greek?

**Vanna** (∂²V/∂F∂σ) measures how delta changes as volatility moves. With both a large spot jump and a large vol spike coinciding, Vanna is non-negligible. Use a central-difference cross-partial:

$$\text{Vanna} = \frac{V(F+h, \sigma+k) - V(F+h, \sigma-k) - V(F-h, \sigma+k) + V(F-h, \sigma-k)}{4 \cdot h \cdot k}$$

with h = `step_S` and k = `step_v`. The Vanna P&L contribution is Vanna × ΔF × Δσ.

Present all results in a `gt::gt()` table. Which position dominates the loss? Is the Greek attribution a reliable approximation here, and why or why not?

```{r q3}

# Compute actual, attributed, unexplained

wtiChange <- 90 - 73
volChange <- .20
timeChange <- stepSizeTheta

changesVec <- c(wtiChange, 0.5 * (wtiChange^2), volChange, timeChange, wtiChange * volChange)

greekNames <- c("Delta", "Gamma", "Vega", "Theta", "Vanna")

# Actual PnL

actualPnL <- port %>%
  dplyr::mutate(priceNew = mapply(priceOption, S = 90, X = K, T2M = (T2M - timeChange), sigma = (sigma + volChange), type = type),
                priceOld = mapply(priceOption, S = F_0, X = K, T2M = T2M, sigma = sigma, type = type)) %>%
  dplyr::mutate(positionValueNew = n * mult * priceNew,
                positionValueOld = n * mult * priceOld) %>%
  dplyr::mutate(actualPnL = positionValueNew - positionValueOld) %>%
  dplyr::select(Position, actualPnL) %>%
  tidyr::pivot_wider(names_from = "Position", values_from = "actualPnL") %>%
  dplyr::mutate(" " = "Actual PnL") %>%
  dplyr::select(c(" ", everything())) %>%
  dplyr::mutate("Portfolio" = sum(across(where(~ is.numeric(.)))))

actAttribTable <- actualPnL

# Comprehensive Table for Greek PnLs

greeks <- portTableFull %>%
  dplyr::filter(underlyingFut == F_0) %>%
  dplyr::mutate(priceUpUp = mapply(priceOption, S = (underlyingFut + stepSizeS), X = K, T2M = T2M, sigma = (sigma + stepSizeVol), type = type),
                priceUpDown = mapply(priceOption, S = (underlyingFut + stepSizeS), X = K, T2M = T2M, sigma = (sigma - stepSizeVol), type = type),
                priceDownUp = mapply(priceOption, S = (underlyingFut - stepSizeS), X = K, T2M = T2M, sigma = (sigma + stepSizeVol), type = type),
                priceDownDown = mapply(priceOption, S = (underlyingFut - stepSizeS), X = K, T2M = T2M, sigma = (sigma - stepSizeVol), type = type)) %>%
  dplyr::mutate(vanna = n * mult * ((priceUpUp - priceUpDown - priceDownUp + priceDownDown) / (4 * stepSizeS * stepSizeVol))) %>%
  dplyr::select(Position, delta, gamma, vega, theta, vanna) %>%
  dplyr::mutate(!!!setNames(map2(.[, -1], changesVec, ~ .x * .y), paste0(greekNames, " PnL"))) %>%
  dplyr::select(Position, contains("PnL"))

# Attributed PnL

attribPnL <- greeks %>%
  dplyr::group_by(Position) %>%
  dplyr::mutate(positionVal = sum(across(where(~ is.numeric(.))))) %>%
  dplyr::ungroup() %>%
  dplyr::select(Position, positionVal) %>%
  tidyr::pivot_wider(names_from = "Position", values_from = "positionVal") %>%
  dplyr::mutate(" " = "Attributed PnL") %>%
  dplyr::select(" ", everything()) %>%
  dplyr::mutate("Portfolio" = sum(across(where( ~is.numeric(.)))))

actAttribTable <- bind_rows(actAttribTable, attribPnL)

# Unexplained PnL

unexplainedPnL <- actAttribTable %>%
  tidyr::pivot_longer(cols = -" ",
                      names_to = "Instrument",
                      values_to = "PnL") %>%
  tidyr::pivot_wider(names_from = " ",
                     values_from = "PnL") %>%
  dplyr::mutate("Unexplained PnL" = `Actual PnL` - `Attributed PnL`) %>%
  dplyr::select(Instrument, "Unexplained PnL") %>%
  tidyr::pivot_wider(names_from = Instrument, values_from = "Unexplained PnL") %>%
  dplyr::mutate(" " = "Unexplained PnL") %>%
  dplyr::select(" ", everything())

actAttribTable <- bind_rows(actAttribTable, unexplainedPnL)

# Greek Attribution Breakdown

attribPnLs <- attribPnL %>%
  dplyr::select(-" ") %>%
  tidyr::pivot_longer(cols = everything(), names_to = "instrument", values_to = "attribPnL") %>%
  dplyr::pull(attribPnL)

attribBreakdown <- greeks %>%
  tidyr::pivot_longer(-Position,
                      names_to = "measure",
                      values_to = "PnL") %>%
  tidyr::pivot_wider(names_from = Position, values_from = PnL) %>%
  dplyr::group_by(measure) %>%
  dplyr::mutate("Portfolio" = sum(across(where(~ is.numeric(.))))) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(!!!setNames(map2(.[, -1], attribPnLs, ~ (.x / .y)), paste0(colnames(.[, -1])))) %>%
  dplyr::mutate(measure = str_replace_all(measure, " PnL", " % of Attribution")
  ) %>%
  dplyr::rename(" " = measure)
  
actAttribTable <- bind_rows(actAttribTable, attribBreakdown)

actAttribTable %>% gt() %>%
  fmt_number(decimals = 2) %>%
  fmt_percent(rows =  c(4:8)) %>%
  tab_header(title = "Actual vs. Attributed PnL") %>%
  tab_style(
    style = cell_borders(sides = "all", color = "black"),
    locations = cells_body()
  )


```

### Analysis of Actual v. Attributed PnL – Iran War

Our short call at a strike of \$80 dominates the loss in this position. This is because that position alone is deep in the money while the other positions are far out of the money. Since we're in a short position, we lose more money as that call goes deeper in the money.

A greek attribution appears to be an unreliable approximation here, due to the sheer magnitude in risk factor shifts. For the overall portfolio, the loss was overstated by \$789,080, which is a consequential amount of estimation error. When looking at the individual positions in the table, their respective attributions are even more off – signalling that the lower unexplained amount for the portfolio may happen to be that way due to luck when summing across positions.

The reason that the approximation breaks down is because of how far the shifts are relative to the values (except theta) at which the local slopes for each risk factor were calculated. Since we have finite number of terms in our Taylor series approximation, which theoretically can contain an infinite number of terms, the error margin grows as we deviate from our local points.

# Question 4 (4 points)

The Iran war scenario (ΔF = +\$17) is an extreme move. How well does the Greek approximation hold as WTI moves further from the current market?

Holding vol and time constant at their initial values, compute Actual P&L (full revaluation) and Attributed P&L (Delta + Gamma only) for the portfolio across ΔF ∈ \[–20, +20\] in \$1 increments.

**a.** Plot Actual, Attributed, and the Residual on the same chart. Mark the Iran war move (ΔF = +17) with a vertical line.

**b.** At what ΔF does the absolute residual first exceed 5% of the absolute Actual P&L? Report in a `gt::gt()` table. What does this tell you about using Greeks for risk management in extreme scenarios?

```{r q4}

lowerBound <- F_0 - 20
upperBound <- F_0 + 20

localDelta <- portTableFull %>%
  dplyr::filter(underlyingFut == F_0) %>%
  dplyr::pull(delta) %>%
  sum(.)

localGamma <- portTableFull %>%
  dplyr::filter(underlyingFut == F_0) %>%
  dplyr::pull(gamma) %>%
  sum(.)

# Comprehensive Table

attribActualCompare <- portTableFull %>%
  dplyr::filter(underlyingFut >= lowerBound & underlyingFut <= upperBound) %>%
  dplyr::select(-c(delta, gamma, vega, theta)) %>%
  dplyr::mutate(positionValue = n * mult * price) %>%
  dplyr::group_by(underlyingFut) %>%
  dplyr::summarise(
    portValue = sum(positionValue)
  ) %>%
  dplyr::ungroup() %>%
  dplyr::mutate("Actual PnL" = portValue - portValueToday) %>%
  dplyr::mutate(delta = localDelta,
                gamma = localGamma) %>%
  dplyr::mutate("Attributed PnL" = (delta * (underlyingFut - F_0)) + (0.5 * gamma * ((underlyingFut - F_0)^2))) %>%
  dplyr::select(underlyingFut, !!"Actual PnL", !!"Attributed PnL") %>%
  dplyr::mutate(underlyingFut = underlyingFut - F_0) %>%
  dplyr::rename("Change in F" = underlyingFut) %>%
  dplyr::mutate("Unexplained PnL" = round(`Actual PnL` - `Attributed PnL`, 2))


# Get % of actual unexplained

percentTable <- attribActualCompare %>%
  dplyr::select(`Change in F`, `Actual PnL`, `Unexplained PnL`) %>%
  dplyr::mutate(`Actual PnL` = abs(`Actual PnL`),
                `Unexplained PnL` = abs(`Unexplained PnL`)) %>%
  dplyr::mutate("% of Actual Unexplained" = `Unexplained PnL` / `Actual PnL`) %>%
  dplyr::filter(`% of Actual Unexplained` >= 0.05) %>%
  dplyr::slice(8:9) %>%
  dplyr::select(`Change in F`, `% of Actual Unexplained`)


```

### Actual, Attributed, & Unexplained PnL across F ∈ \[–20, +20\]

```{r, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}

attribActualCompare %>%
  plotly::plot_ly() %>%
  plotly::add_trace(
    x = ~ `Change in F`,
    y = ~ `Actual PnL`,
    type = "scatter",
    mode = "lines",
    name = "Actual PnL"
  ) %>%
  plotly::add_trace(
    x = ~ `Change in F`,
    y = ~ `Attributed PnL`,
    type = "scatter",
    mode = "lines",
    name = "Attributed PnL"
  ) %>%
  plotly::add_trace(
    x = ~ `Change in F`,
    y = ~ `Unexplained PnL`,
    type = "scatter",
    mode = "lines",
    name = "Unexplained PnL"
  ) %>%
  plotly::layout(shapes = list(
    list(
      type = "line", 
      y0 = min(attribActualCompare$`Actual PnL`), 
      y1 = max(attribActualCompare$`Actual PnL`), 
      x0 = 17, 
      x1 = 17,
      line = list(color = "darkred"))),
    title = list(text = "Actual, Attributed, & Unexplained PnL across F ∈ [–20, +20]"),
    xaxis = list(title = "ΔF ($)"),
    yaxis = list(title = "PnL ($)"))

```

\*\* The red vertical line above showcases the Iran war scenario \*\*

```{r, eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}

percentTable %>%
  gt() %>%
  tab_header(title = "5% Breaking Point for Residual over Actual PnL") %>%
  tab_style(
    style = cell_borders(sides = "all", color = "black"),
    locations = cells_body()
  ) %>%
  fmt_number(columns = 2, decimals = 2) %>%
  fmt_percent(columns = 2, rows = c(1:2))

```

The table above showcases the "breaking point" ΔF after which the the absolute residual exceeds 5% of the absolute actual PnL. Beyond these points (in magnitude in their respective directions), that percentage only grows. This is consistent with the plot generated above, where we can see the attributed PnL series begin to deviate from the actual PnL series after a certain price change threshold.

In the context of using greeks for risk management in extreme scenarios, the figures above illustrate an important limitation – the larger the change in risk factor, the less accurately greeks will explain value movements. For a deeper dive into this, see "Key Learnings."

# Question 5 (6 points)

Discuss the key learnings and insights from a Greek risk management perspective. Your explanation must:

-   Interpret the **full revaluation MTM profile** from Q1(d): describe the portfolio's curvature structure across the \$50–\$100 range. Identify the regions of maximum and minimum Gamma. Explain what the changing Gamma means for a desk running this portfolio — specifically, how frequently delta rebalancing is needed as WTI moves through the strikes.
-   Interpret the **full revaluation P&L and Gamma profiles** from Q1(d): describe the portfolio's curvature structure and identify which positions drive positive vs negative Gamma regions.
-   Identify which position and which Greek drove the largest loss in the Iran war scenario, and explain why that was predictable from the portfolio structure.
-   Explain the role of **Vanna**: why is it non-trivial when both a large spot jump and a large vol spike occur simultaneously?
-   Explain why the Greek approximation (Q4) breaks down at extreme moves, and what this means for a real desk managing this book.

Pay particular attention to why textbook delta-hedging assumptions break down in geopolitical gap scenarios, and what risk tools a real energy desk would use instead.

### Key Learnings:

-   The full revaluation MTM profile showcased a downward sloping, slightly concave curve – signalling an overall negative delta position, meaning the portfolio loses value as the price of WTI increases. The magnitude of the portfolio gamma was highest near the strike of the short put (55) and lowest near the strike of the long put (70). Gamma magnitude decreased the price of WTI futures approached the strike of the long put. The closer the value of the underlying is to the strike of short positions, the more frequently a delta-hedger would need to rebalance.

-   The portfolio was slightly concave – meaning that gamma was smaller at WTI prices closer to the strike of the long put (70) and higher for WTI prices closer to the strikes of the short put & call. After strike 80, the gamma begins decreasing again as it does not approach any more strike prices. This is consistent with the gamma profile created. The overall portfolio is consistently negative gamma, which is driven by the short positions in the put at 55 and call at 80. The long put offsets those two positions, which is why we see the gamma magnitude decreasing at WTI prices closer to the long put's strike.

-   The position that drove the largest loss in the Iran war scenario was the short call at strike 80. This is easily explained by the portfolio structure – the short call is the only instrument at that point which is in the money. This causes the value of that position to skyrocket, and since we're short it means that we lose a lot of money.

-   Vanna is the rate of change of delta with respect to the change in volatility. Delta and volatility interact because delta can be seen as a rough estimate of the probability that an option expires in the money, according to a risk-neutral framework. Since volatility is essentially how "wide" a distribution of prices are, it increases the probability that an option expires in the money – therefore increasing delta.

-   A greek approximation breaks down at extreme moves because greeks are only reliable at approximating PnL when moves are small and close to the expansion point. Mathematically, we are unable to fully explain PnL for large shifts because of the lack of higher-order terms at our disposal. This doesn't dismiss the effectiveness of greek approximations – rather it showcases that it's only one aspect of risk management. A real desk managing this book can use greeks for smaller, more reasonable risk factor shifts and use other tools – such as full revaluation scenario analysis, analyzing tail risk, etc. – for investigating extreme moves.
